Burning sage: Reversing the curse of dimensionality in the visualisation of high-dimensional data

### 2020-09-22 paper/paper.Rmd
fig 2 explanation

---
title: |
  Burning sage: Reversing the curse of dimensionality in the visualisation of high-dimensional data
type: Article
author:
  - name: Ursula Laa
    affil: a, b
    email: ursula.laa@monash.edu
  - name: Dianne Cook
    affil: b
    email: dicook@monash.edu
  - name: Stuart Lee
    affil: b, c
    email: stuart.lee1@monash.edu
affiliation:
  - num: a
    address: |
      School of Physics and Astronomy, Monash University
  - num: b
    address: |
      Department of Econometrics and Business Statistics, Monash University
  - num: c
    address: |
      Molecular Medicine Division, Walter and Eliza Hall Institute, Parkville, Australia
bibliography: biblio.bib
geometry: margin=2.5cm
abstract: |
  In high-dimensional data analysis the curse of dimensionality reasons that points tend to be far away from the center of the distribution and on the edge of high-dimensional space. Contrary to this, is that projected data tends to clump at the center. This gives a sense that any structure near the center of the projection is obscured, whether this is true or not. A transformation to reverse the curse, is defined in this paper, which uses radial transformations on the projected data. It is integrated seamlessly into the grand tour algorithm, and we have called it a burning sage tour, to indicate that it reverses the curse. The work is implemented into the tourr package in R. Several case studies are included that show how the sage visualisations enhance exploratory clustering and classification problems.
keywords: |
  data visualisation; grand tour; statistical computing; statistical graphics; multivariate data; dynamic graphics
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
output: rticles::tf_article
keep_tex: true
---

@@

To understand why points tend to be away from the center in the high-dimensional space, but pile near the center in low-dimensional projections, we need to consider the projected volume. It is useful to think of the data as being distributed in a hypersphere, i.e. all data points are within a specified distance from the center. Compared to the common assumption of a distribution in a box or hypercube this has the advantage of being rotation invariant, and more accurately captures commonly observed distributions.

```{r, out.width="60%", fig.cap="The diagram on the left is illustrating the different volumes considered for a three-dimensional sphere. The full sphere has volume $V(R, p)$. Within a radius $r$ the sphere contains the reduced volume $V(r; p, R)$, shown in blue, but the projected volume within a radius $r$ in a two-dimensional plane is much larger, given by the volume of the cylinder with rounded caps, $V_{2D}(r; p ,R)$, shown in red. The intersection of the plane with the sphere is illustrated in grey, and the plane with both radii is shown in the right diagram.", fig.align = "center"}

@@

We can picture this for a 3D sphere. In any 2D projection a large fraction of the volume will be projected onto a small region near the center of the resulting disk. This is a consequence of the curved shape, which extends further in the orthogonal direction at the center of the disk, since the projection corresponds to an integration over the orthogonal space. The effect will be further enhanced when projecting onto 1D instead. The same intuition holds as we increase the number of original dimensions that are projected down to one or two dimensions for visualization, with increasing fractions of the volume being projected onto a small area around the center, see Figure \ref{fig:density} for illustration.

To reverse this effect, we introduce a radial transformation that redistributes the projected points, such that equal volume in the original ($p$-dimensional) space is projected onto equal area in a 2-dimensional projection. Note that it is straightforward to generalize this for $d$-dimensional projections by mapping onto equal $d$-dimensional volume instead.

>>>>

---
title: |
  Burning sage: Reversing the curse of dimensionality in the visualisation of high-dimensional data
type: Article
author:
  - name: Ursula Laa
    affil: a, b
    email: ursula.laa@monash.edu
  - name: Dianne Cook
    affil: b
    email: dicook@monash.edu
  - name: Stuart Lee
    affil: b, c
    email: stuart.lee1@monash.edu
affiliation:
  - num: a
    address: |
      School of Physics and Astronomy, Monash University
  - num: b
    address: |
      Department of Econometrics and Business Statistics, Monash University
  - num: c
    address: |
      Molecular Medicine Division, Walter and Eliza Hall Institute, Parkville, Australia
bibliography: biblio.bib
geometry: margin=2.5cm
abstract: |
  In high-dimensional data analysis the curse of dimensionality reasons that points tend to be far away from the center of the distribution and on the edge of high-dimensional space. Contrary to this, is that projected data tends to clump at the center. This gives a sense that any structure near the center of the projection is obscured, whether this is true or not. A transformation to reverse the curse, is defined in this paper, which uses radial transformations on the projected data. It is integrated seamlessly into the grand tour algorithm, and we have called it a burning sage tour, to indicate that it reverses the curse. The work is implemented into the tourr package in R. Several case studies are included that show how the sage visualisations enhance exploratory clustering and classification problems.
keywords: |
  data visualisation; grand tour; statistical computing; statistical graphics; multivariate data; dynamic graphics; data science; machine learning
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
output: rticles::tf_article
keep_tex: true
---

@@

To understand why points tend to be away from the center in the high-dimensional space, but crowd the center in low-dimensional projections, it is helpful to consider the projected volume relative to high-dimensional volume. To avoid edge effects and to impose rotation invariance, we will start from the data being uniformly distributed in a hypersphere, i.e. all data points are within a specified distance from the center. This makes calculations more tractable than assuming a uniform distribution in hypercube (box).

```{r, out.width="60%", fig.cap="Illustration (and notation) for describing the element used in the burning sage transformation. The 3D sphere (left) shows the different volumes to be compared. The full sphere has volume $V(R, p)$. Within a radius $r$ the sphere contains the reduced volume $V(r; p, R)$, shown in blue, but the projected volume within a radius $r$ in a two-dimensional plane is much larger, given by the volume of the cylinder with rounded caps, $V_{2D}(r; p ,R)$, shown in red. The intersection of the plane with the sphere is illustrated in grey, and the plane representing the projection with both radii is shown at right.", fig.align = "center"}

@@

Figure \ref{fig:density} illustrates the comparison to be made, using something we can easily picture, a 3D sphere. Projecting the data from within a 3D sphere to 2D (grey disk) will result in mass being condensed into the disk. Imagine comparing the volume of a cylinder at different locations in the disk. A centered cylinder has more volume. This is exaggerated as $p$ increases: the centered cylinder has much more volume than any other cylinder.

To reverse this effect, we introduce a radial transformation that redistributes the projected points, such that equal volume in the original ($p$-dimensional) space is projected onto equal area in a 2-dimensional projection. Note that this can be generalized for $d$-dimensional projections by mapping onto equal $d$-dimensional volume instead.

### 2020-09-22 paper/paper.tex
fig 2 explanation

\begin{keywords}
data visualisation; grand tour; statistical computing; statistical
graphics; multivariate data; dynamic graphics
\end{keywords}

@@

To understand why points tend to be away from the center in the
high-dimensional space, but pile near the center in low-dimensional
projections, we need to consider the projected volume. It is useful to
think of the data as being distributed in a hypersphere, i.e.~all data
points are within a specified distance from the center. Compared to the
common assumption of a distribution in a box or hypercube this has the
advantage of being rotation invariant, and more accurately captures
commonly observed distributions.

@@

\caption{The diagram on the left is illustrating the different volumes considered for a three-dimensional sphere. The full sphere has volume $V(R, p)$. Within a radius $r$ the sphere contains the reduced volume $V(r; p, R)$, shown in blue, but the projected volume within a radius $r$ in a two-dimensional plane is much larger, given by the volume of the cylinder with rounded caps, $V_{2D}(r; p ,R)$, shown in red. The intersection of the plane with the sphere is illustrated in grey, and the plane with both radii is shown in the right diagram.}\label{fig:unnamed-chunk-1}
\end{figure}

We can picture this for a 3D sphere. In any 2D projection a large
fraction of the volume will be projected onto a small region near the
center of the resulting disk. This is a consequence of the curved shape,
which extends further in the orthogonal direction at the center of the
disk, since the projection corresponds to an integration over the
orthogonal space. The effect will be further enhanced when projecting
onto 1D instead. The same intuition holds as we increase the number of
original dimensions that are projected down to one or two dimensions for
visualization, with increasing fractions of the volume being projected
onto a small area around the center, see Figure \ref{fig:density} for
illustration.

To reverse this effect, we introduce a radial transformation that
redistributes the projected points, such that equal volume in the
original (\(p\)-dimensional) space is projected onto equal area in a
2-dimensional projection. Note that it is straightforward to generalize
this for \(d\)-dimensional projections by mapping onto equal
\(d\)-dimensional volume instead.

>>>>

\begin{keywords}
data visualisation; grand tour; statistical computing; statistical
graphics; multivariate data; dynamic graphics; data science; machine
learning
\end{keywords}

@@

To understand why points tend to be away from the center in the
high-dimensional space, but crowd the center in low-dimensional
projections, it is helpful to consider the projected volume relative to
high-dimensional volume. To avoid edge effects and to impose rotation
invariance, we will start from the data being uniformly distributed in a
hypersphere, i.e.~all data points are within a specified distance from
the center. This makes calculations more tractable than assuming a
uniform distribution in hypercube (box).

@@

\caption{Illustration (and notation) for describing the element used in the burning sage transformation. The 3D sphere (left) shows the different volumes to be compared. The full sphere has volume $V(R, p)$. Within a radius $r$ the sphere contains the reduced volume $V(r; p, R)$, shown in blue, but the projected volume within a radius $r$ in a two-dimensional plane is much larger, given by the volume of the cylinder with rounded caps, $V_{2D}(r; p ,R)$, shown in red. The intersection of the plane with the sphere is illustrated in grey, and the plane representing the projection with both radii is shown at right.}\label{fig:unnamed-chunk-1}
\end{figure}

Figure \ref{fig:density} illustrates the comparison to be made, using
something we can easily picture, a 3D sphere. Projecting the data from
within a 3D sphere to 2D (grey disk) will result in mass being condensed
into the disk. Imagine comparing the volume of a cylinder at different
locations in the disk. A centered cylinder has more volume. This is
exaggerated as \(p\) increases: the centered cylinder has much more
volume than any other cylinder.

To reverse this effect, we introduce a radial transformation that
redistributes the projected points, such that equal volume in the
original (\(p\)-dimensional) space is projected onto equal area in a
2-dimensional projection. Note that this can be generalized for
\(d\)-dimensional projections by mapping onto equal \(d\)-dimensional
volume instead.
